\Large \bf \Toolname: Cache Line Contention Detection for COTS Binaries

### 2014-12-15 HotOS'15/paper.tex
Working on the methodology paper. Having some issues on what the right amount of detail is for describing the implementation of ownership tracking. There are many valid designs, and there is a single correct design. The correct design is heavyweight, requiring a CMPXCHG. I chose one of the valid designs that seemed to have a reasonable state transition diagram.

%--------sigalternate----------
\documentclass{sig-alternate}
%--------/sigalternate----------

%\documentclass[letterpaper,twocolumn,10pt]{article}
%\usepackage{endnotes,multirow}
\usepackage{refstyle,amsmath,chngcntr}
\usepackage{epsfig,subfigure,framed}
\usepackage{textcomp} % for tilde
\usepackage{paralist} % for in-paragraph lists
\usepackage{graphicx}
\usepackage{hyperref}

@@

@@

%--------sigalternate----------\titlespacing\section{0pt}{5pt plus 4pt minus 2pt}{5pt plus 2pt minus 2pt}        
%--------sigalternate----------\titlespacing\subsection{0pt}{5pt plus 4pt minus 2pt}{5pt plus 2pt minus 2pt}

@@

% For leaving some comments in the draft.
\newcommand{\comment}[1]{}
\newcommand{\TextToolname}{CatcheScan}
\newcommand{\Toolname}{\textsc{\TextToolname{}}}
\newcommand{\BoldToolname}{\textsc{\bf\TextToolname{}}}

@@

@@

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf \Toolname: Cache Line Contention Detection for COTS Binaries}
\numberofauthors{3}
\author{
% 1st. author
\alignauthor
Peter Goodman\\
       \email{pag@cs.toronto.edu}
\and
% 2nd. author
\alignauthor
Ashvin Goel\\
       \email{asvhin@eecg.toronto.edu}
% 3rd. author
\alignauthor
Angela Demke Brown\\
       \email{demke@cs.toronto.edu}
}
\maketitle

@@

@@

\Toolname{} uses dynamic binary translation (DBT) to instrument a program binary and detect cache line contention
at runtime. Our approach combines unlimited watchpoints \cite{UnlimitedWatchpoints,DynamoRIOWatchpoints}, data
sampling, and a lightweight cache line ownership tracking mechanism to detect cache line contention. The key insight
of our approach is that if we know \begin{inparaenum}[i)]
\item what code is executed by more than one thread,
\item what types of data are accessed by more than one thread, and
\item what types of data are accessed by what code, then
\end{inparaenum}
we can efficiently detect cache line contention by selectively instrumenting only the code and the data that matters.

@@

\paragraph{Contributions}
This paper makes the following significant contributions:
\begin{itemize}
\item It describes \Toolname{}, a software-only system that dynamically and precisely diagnoses cache line
contention in multi-threaded, COTS binaries. This enables \Toolname{} to detect performance
and correctness bugs in programs using home-grown synchronization primitives and lock-free algorithms.

@@

%from wasting cycles looking for contention on
%unshared data

%This is beneficial because it improves coverage: if 90\% of false sharing occurs because of accesses
%to only 10\% of the data, then \Toolname{} will not miss those races by blindly focusing on the other 90\% of data

\item It describes a novel separation of cache line contention detection into two phases: \begin{enumerate}
\item[1) \emph{Training:}] an ahead-of-time, heavyweight analysis of the program's execution that
discovers what code can contain cache line contention, and what data is shared by multiple threads. A key
aspect of this phase is that it \emph{does not} need to exhibit any cache line contention; the only requirement
is that training and production workloads exercise/cover the code in similar ways.

@@

\item[2) \emph{Detection:}] an online analysis that detects cache line contention by selectively
instrumenting only the relevant code and sampling only the relevant data as recognized by the
offline training phase.
\end{enumerate}

@@

%It describes a novel separation of cache line contention detection into two phases: \begin{enumerate}
%\item[1) \emph{Training:}] This phase uses heavyweight instrumentation to discover \begin{inparaenum}
%\item what code is hot;
%\item what code is executed by more than one thread;
%\item what code accesses what data; and,
%\item function call graphs.
%\end{inparaenum}
%
%This information enables the detection phase to selectively prune the set of code that is instrumented,
%and the set of data that is sampled in order to reduce overheads. 
%
%
%
%\item[2) \emph{Detection:}] b
%\end{enumerate}

@@

%\begin{enumerate}
%\item Detecting what code is important to the 
%\end{enumerate}
%
%analyzing
%the 
%
%the detection
%of what code should be instrumented and what data should be sampled is \emph{not} time-sensitive, and
%therefore can be performed any time. 
%
%using results gathered from
%an offline learning and analysis phase. This separation is motivated by recognizing that

@@

\Toolname{} uses DBT to instrument the memory accesses of a program while it is running. During the offline
learning phase, DBT is used to figure out what code is executed by more than one thread, what data is accessed
by more than one thread, and what code accesses what data. During its online detection phase, \Toolname{} uses
an implementation of unlimited watchpoints watch for accesses to a representative sample of cache lines, assign
owning threads to those cache lines, and finally detect transfers in ownership.

\subsection{Definitions}
\Toolname{} looks for cache line contention bugs of the following forms:
\begin{description}
\item[Data races] Data races occur when the same program variable is concurrently modified by two threads, or 
modified by one thread and concurrently accessed by another thread, where neither access is atomic (e.g. using
and atomic read-modify-write instruction).

\item[True sharing] True sharing has the same access pattern as a data race, except that one or both operations
are atomic.

\item[False sharing] False sharing occurs when two semantically related variables (e.g. two fields in a struct) occupy
the same cache line and are concurrently accessed, thus introducing contention on the cache line.

\item[Pseudo sharing] Pseudo sharing occurs when two semantically unrelated variables (e.g. fields in adjacent structs,
adjacent array entries) occupy the same cache line and are concurrently accessed \cite{DegenerateSharingAndFalseCoherence}.
\end{description}

%\subsection{Mechanisms}

@@

%% --------------------------------------
\paragraph{DataCollider}
DataCollider \cite{DataCollider} is a precise, sampling-based race detector for the WIndows kernel. DataCollider's key
novelties are its extremely low overhead, as well as its generality: it detects data races without any understanding of
a system's underlying synchronization primitives. DataCollider uniformly samples code by introducing breakpoints into the code.
When a breakpoint is hit, a hardware watchpoint is added to data that was about to be accessed by the interrupted instruction.
Then, the interrupted thread is paused for a short period of time to give another thread the opportunity to access the
same data. If the data is accessed by a second thread while the first thread is paused, then the watchpoint will detect the
access, thus allowing DataCollider to detect a race ``in the act.""

While this approach is elegant in its simplicity, it has two major drawbacks. First, its dependence on hardware watchpoints
limits it to being able to detect races on at most four distinct 8-byte memory locations simultaneously \cite{IntelSDManual}.
Second, it cannot detect guarded cache line line conflicts: if the same cache line is concurrently accessed by two threads, but
the accesses are guarded by a lock, then the pausing introduced by the first thread will prevent the watchpoint from being hit
in the second thread (because the second thread will block on the lock).

@@

Intel's VTune \cite{IntelVTune} an an imprecise, sampling-based profiler that can detect false sharing using hardware performance
counters and monitoring events. It provides detailed, architecture-specific performance-tuning suggestions. Unlike \Toolname{},
VTune can only report individual program counters that general cache coherency events, but cannot correlate threads by showing
two contending stack traces.

@@

\paragraph{Plastic}
Plastic \cite{Plastic} is a Xen-based dynamic false sharing detector and repairer. Like \Toolname{}, it takes a pipeline-based
approach that successively narrows down on the root causes of false sharing. Plastic also uses DBT as part of its
repair mechanism. Plastic's pipeline begins by using hardware performance counters related to cache contention
events to detect the presence of false-sharing. Then, it applies a similar technique to MultiRace \cite{MultiRace}, in
that it tracks the dominant access pattern by each thread to memory on a shared page. This information informs
their repair mechanism, which transparently re-maps memory accesses at a byte granularity.

\paragraph{SHERIFF}
SHERIFF \cite{SHERIFF} is a false-sharing detector and repairer that works in a similar way to MultiRace and Plastic. Initially,
it applies read-only hardware page protection to all shared memory. Then, as threads attempt to write to protected pages, those
pages are privately mapped as read/write. When the same page is privately mapped by two or more threads, SHERIFF applies
the twinning and diffing approach of TreadMarks \cite{TreadMarks} to detect local changes to shared pages and narrow down
on instances of false sharing. Like \Toolname{}, SHERIFF uses a coarse-grained mechanism to hint at the potential for races.

@@

%Under this categorization scheme, \Toolname{} is a dynamic, precise, variable-granularity, mixed-mode
%false-sharing detector. It draws inspiration from a number of existing race detectors, including LiteRace
%\cite{LiteRace}, MultiRace \cite{MultiRace}, RaceMob \cite{RaceMob}, RaceTrack \cite{RaceTrack}, and
%DataCollider \cite{DataCollider}.

LiteRace \cite{LiteRace} is a precise, sampling-based dynamic race detector implemented on top of the Microsoft
Pheonix binary translator. It samples at a function granularity, and executes either one specialized version of the
code or another. The sampling rate for a function is inversely proportional to how often it executes. This is based
on the insight that races are likely to occur in infrequently executed code, because otherwise they are benign or
would have been noticed and resolved. This insight, called the ``cold code hypothesis", motivated \Toolname's
post-mortem ranking scheme: cases of false sharing are more highly ranked when one---but \emph{not} both---of
the participating threads are executing cold code. This variant on the cold code hypothesis emphasizes the similarly
hard-to-find case of cold code pessimizing hot code.

\newcommand\TypeIdPair{$\langle PC,size \rangle$ }

MultiRace \cite{MultiRace} is a comprehensive, dynamic race detector that employs a hybrid happens-before and
lockset-based approach. It operates at an object granularity, and therefore reports false-positives. It is unique
in that it uses ``pointer swizzing" as a means of remapping memory to ``minipages" that track the dominating
access pattern of a thread (read-write > read-only > no-access). \Toolname{} uses a similar pointer swizzing approach, but
in the form of address watchpoints \cite{AddressWatchpoints}. Pointers to heap-allocated objects are ``swizzled" to
include a type identifier that is unique to a \TypeIdPair{} pair. This improves the diagnosis of false-sharing,
because it allows \Toolname{} to distinguish between two threads accessing different fields of the same object, and
two threads accessing two nearby objects of the same (or different) types.

RaceMob \cite{RaceMob} and RaceTrack \cite{RaceTrack} are two hybrid happens-before and lockset-based race detectors.
RaceMob begins by statically analyzing source code to find potential races (lockset), then dynamically verifies those races
(happens-before). RaceTrack is a JIT-based race-detector for Microsoft's common language runtime (CLR). It uses the
concept of threadsets to efficiently maintain happens-before relationships for memory locations and detect the possible
presence of a race and uses lockets to confirm predicted races. The novelty of RaceTrack is that it adapts the detection
granularity from object-granularity to field-granularity. \Toolname{} follows the same high-level approach as both RaceMob
and RaceTrack in that it attempts to predict the source of false-sharing at one granularity, then verify those predictions 
at another granularity.

@@

%\paragraph{Sampling-based}

%\paragraph{Comprehensive}

@@

This section introduces our approach to finding false sharing bugs. The goal of our approach is extract beliefs about
where false sharing might be, then iteratively strenghten those beliefs, either verifying or discarding them in the process.

We begin with the same intuition as SHERIFF: if false sharing occurs on a single cache line, then that implies that two
threads are simultaneously accessing the same page of memory \cite{SHERIFF}. Therefore, by detecting contention on
a single page, we can establish a belief that false sharing might occur on the contended page.

Unfortunately, a page-granularity detection of false sharing is na{\"i}ve: it provides no guidance as to the location of the
contended cache line, if any. Our insight is that we can narrow down on instances of false sharing by varying the
granularity of the mechanism used to detect thread contention. That is, we can start with a coarse-grained but
innacurate approach that tells us what code might cause false sharing, thus establishing an initial belief set. Then, we
can strengthen or discard beliefs by repeating this same process, but using finer-grained detection mechanisms that
apply only to code thought to contain false sharing.

The remainder of this section explains how we proxy memory access behavior (\Secref{proxymem}), how we use sampling
to detect contending accesses to shared memory (\Secref{heapsample}), how we reduce false negatives with our
sampling-based approach (\Secref{falsenegs}), how we achieve precision by eliminating false positives (\Secref{falsepos}),
and finally how we rank discovered instances of false sharing (\Secref{ranking})

% --------------------------------------
\subsection{Proxy Memory and Tagged Reads}\label{sec:proxymem}
Proxy memory is a form of variable-granularity shadow memory that proxies actual memory access behavior. The key
idea is to instrument all loads and stores in such a way that their load/store behavior is reflected in the shadow
memory in a way that can be externally observed. For example, one way to achieve this is to instrument every load
to read from shadow memory, every store to modify shadow memory, and trap loads and stores by adding hardware
watchpoints to shadow memory locations. Such heavyweight instrumentation would be inefficient, and so we apply
two optimizations: belief-based selective instrumentation, and tagged reads.

\paragraph{Selective Instrumentation}
The first optimization to proxy memory is to not instrument all code to operate on proxy memory. Initially, \Toolname{}
comprehensively instruments all code, thus incurring significant overheads. However, as beliefs about false sharing
are refined, less and less code is instrumented, thus allowing code to run faster when precision during identification of
potential false sharing is needed.

\paragraph{Tagged Reads}
The second optimization avoids the problem of introducing unnecessary cache coherency traffic by treating proxy memory
as read-only, and disambiguating the intention of an operation (load, store) by using a \emph{tagged read}. In our implementation,
a read operations (on the proxy memory) that intend to communicate loads (of actual memory) are distinguished from
read operations that intend to communicate a store by using different encodings of the logical \texttt{OR} instruction.

% --------------------------------------
\subsection{Heap Sampling}\label{sec:heapsample}

>>>>

\documentclass[letterpaper,twocolumn,10pt]{article}

\usepackage[margin=1in]{geometry}
%\usepackage{endnotes,multirow}
\usepackage{refstyle,amsmath,chngcntr}
\usepackage{epsfig,subfigure,framed}
\usepackage{textcomp} % for tilde
\usepackage{paralist} % for in-paragraph lists
\usepackage{parskip} % For making paragraphs split by new lines instead of indents
\usepackage{graphicx}
\usepackage{hyperref}

% Used for changing the spacing after the title, sections, and subsections.
\usepackage{titlesec,titling}

@@

\lstdefinelanguage
   [x64]{Assembler}     % add a "x64" dialect of Assembler
   [x86masm]{Assembler} % based on the "x86masm" dialect
   % with these extra keywords:
   {morekeywords={LOCK,XCHG,CMP,JZ,JNZ,CALL, %
                  rax,rdx,rcx,rbx,rsi,rdi,rsp,rbp, %
                  r8,r8d,r8w,r8b,r9,r9d,r9w,r9b}} % etc.

@@

@@

% For leaving some comments in the draft.
\newcommand{\comment}[1]{}
\newcommand{\TextToolname}{Malcontent}
\newcommand{\Toolname}{\textsc{\TextToolname{}}}
\newcommand{\BoldToolname}{\textsc{\bf\TextToolname{}}}

@@

% Modify spacing before the title
\setlength{\droptitle}{-3em}

@@

%don't want date printed
\date{}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf \Toolname: Cache Line Contention Detection for COTS Binaries}
\author{
{\rm Peter Goodman} \hspace{1.5em} {\rm Angela Demke Brown} \hspace{1.5em} {\rm Ashvin Goel}\\
University of Toronto
} % end author
\maketitle

@@

\subsection*{Abstract}
Cache-line contention is an increasingly important problem in multi-threaded applications running on cache coherent SMPs.
Contention on cache lines is symptomatic of both performance and correctness bugs. In terms of performance, cache line
contention limits scalability by introducing extra latency when accessing shared memory.  In terms of correctness, cache
line contention can also be caused by data races, which introduce non-determinism into one's program.

In this paper, we describe \Toolname, a system that detects and diagnoses cache line contention in arbitrary binary programs.
\Toolname{} refines existing contention detection approaches in several novel ways. First, we describe a method training
a detection system to focus on only the code and data that matter to the problem. Second, we describe a new form of
data-centric sampling: comprehensive data sampling. Our sampling approach uses unlimited watchpoints to uniformly
sample memory locations according to the type of data stored in memory. This avoids sampling bias where a large portion
of a program's memory is unshared.

@@

\Toolname{} uses dynamic binary translation (DBT) to instrument a program binary and detect cache line contention
at runtime. Our approach combines unlimited watchpoints \cite{UnlimitedWatchpoints,DynamoRIOWatchpoints}, data
sampling, and a lightweight cache line ownership tracking mechanism to detect cache line contention. The key insight
of our approach is that if we know \begin{inparaenum}[i)]
\item what types of data are accessed by more than one thread, and
\item what types of data are accessed by what code, then
\end{inparaenum}
we can efficiently detect cache line contention by selectively instrumenting only the code and the data that matters.

@@

% --------------------------------------
\subsection{Contributions}
This paper makes the following significant contributions:
\begin{itemize}
\item It describes \Toolname{}, a software-only system that dynamically and precisely diagnoses cache line
contention in multi-threaded, COTS binaries. This enables \Toolname{} to detect performance
and correctness bugs in programs using home-grown synchronization primitives and lock-free algorithms.

@@

%from wasting cycles looking for contention on
%unshared data
%This is beneficial because it improves coverage: if 90\% of false sharing occurs because of accesses
%to only 10\% of the data, then \Toolname{} will not miss those races by blindly focusing on the other 90\% of data

\item It describes a novel separation of cache line contention detection into two phases:

1. {\bf Training:} an ahead-of-time, heavyweight analysis of the program's execution that
discovers what code can contain cache line contention, and what data is shared by multiple threads. A key
aspect of this phase is that it \emph{does not} need to exhibit any cache line contention; the only requirement
is that training and production workloads exercise/cover the code in similar ways.

@@

2. {\bf Detection:} an online analysis that detects cache line contention by selectively
instrumenting only the relevant code and sampling only the relevant data as recognized by the
offline training phase.

@@

%It describes a novel separation of cache line contention detection into two phases: \begin{enumerate}
%\item[1) \emph{Training:}] This phase uses heavyweight instrumentation to discover \begin{inparaenum}
%\item what code is hot;
%\item what code is executed by more than one thread;
%\item what code accesses what data; and,
%\item function call graphs.
%\end{inparaenum}
%
%This information enables the detection phase to selectively prune the set of code that is instrumented,
%and the set of data that is sampled in order to reduce overheads. 
%
%\item[2) \emph{Detection:}] 
%\end{enumerate}

@@

%the detection of what code should be instrumented and what data should be sampled is \emph{not} time-sensitive, and
%therefore can be performed any time. 
%
%using results gathered from an offline learning and analysis phase. This separation is motivated by recognizing that

@@

\Toolname{} uses the Granary DBT \cite{Granary} framework to instrument the memory accesses of a program while
it is running. During the offline training phase, DBT is used to figure out what data is accessed by more than one thread,
and what code accesses that data. During its online detection phase, \Toolname{} uses Granary's implementation of unlimited
watchpoints to watch for accesses to a representative sample of cache lines, assign owning threads to those cache lines,
and finally detect transfers in ownership.

\subsection{Definitions}
\Toolname{} looks for cache line contention bugs of the following forms:

{\bf Data races} occur when the same program variable is concurrently modified by two threads, or 
modified by one thread and concurrently accessed by another thread, and where neither access is atomic (e.g. using
and atomic read-modify-write instruction).

{\bf True sharing} has the same access pattern as a data race, except that one or both operations
are atomic.

{\bf False sharing} occurs when two semantically related variables (e.g. two fields in a struct) occupy
the same cache line and are concurrently accessed, thus introducing contention on the cache line.

{\bf Pseudo sharing} occurs when two semantically unrelated variables (e.g. fields in adjacent structs,
adjacent array entries) occupy the same cache line and are concurrently accessed \cite{DegenerateSharingAndFalseCoherence}.

% --------------------------------------
\subsection{Mechanisms}

% --------------------------------------
\paragraph{Training}
%\paragraph{Code Concurrency Analysis }
%\Toolname{} analyzes code to determine what code is executed by more than one thread. This analysis allows \Toolname{}
%to either not instrument, or only very lightly instrument, non-concurrent code during its detection phase. The insight is
%that code that does get executed by more than one thread is more likely to operate on shared data than code that is
%only ever observed (during training) to have been executed by a single thread.
%
%There are some important cases where this insight does not hold. For example, in a single writer, multiple reader queue,
%the code implementing writes to the queue might only be executed by a designated writer thread, and therefore never
%be marked as concurrent. Obviously, there is contention bet

%\paragraph{Data Concurrency Analysis}
\Toolname's training phase focuses on reducing detection overheads by finding code that doesn't need to be instrumented
during the detection phase. The key challenge of this phase is determining what code can safely be pruned without negatively
affecting the accuracy of the detection phase. Our insight is that the detection phase should only look for cache contention
on accesses to shared data structures. To do this, we must know \begin{inparaenum}
\item what data structures are shared, and
\item what code accesses those data structures.
\end{inparaenum}

\newcommand\TypeId{$TypeID$}
\newcommand\TypeIdPair{$\langle PC,size \rangle$ }

We implemented these analyses using Granary's address watchpoints \cite{AddressWatchpoints} system, which allows us to
``taint" addresses and then interpose on instructions using tainted addresses to access memory. Using Granary, we interpose on
all dynamic memory allocator functions and taint every address returned by those functions with an abstract \TypeId{}.
Each \TypeId{} uniquely identifies an allocation context in terms of a \TypeIdPair{} pair, where $PC$ is the program counter
to which the allocation function will return, and $size$ is the number of bytes being allocated. This arrangement allows us to
approximate the actual number of distinct types of program data structures for arbitrary binaries, even in the absence of
static source code information or debug information.

We use per-type shadow memory to detect data structure sharing accross multiple threads.
We associate each \TypeId{} with a small amount of shadow memory: $\langle InitialThreadId,\linebreak[0]IsShared \rangle$. This
arrangement means that every allocated address tainted by the same \TypeId{} is indirectly associated with the same
shadow memory. If this is the first allocation for a \TypeId, then the shadow memory is initialized to
$\langle CurrThreadId,\linebreak[0]false \rangle$. Sharing is detected and marked either by subsequent allocations for the
same \TypeId{} but by different threads than the initial thread, or by accesses to tainted addresses by threads other
than the initial thread.

To detect what code accesses what types of data, we propagate the {\TypeId}s to code on every memory access to
a tainted address. Finally, to determine what code accesses shared data, we combine the output of both analyses to
produce, for each basic block of code executed during training, we report on whether or not it contained accesses
to shared data structures. During the detection phase, we heavily instrument only those blocks that contain accesses
to shared data structures, or (to be conservative) were never executed during the training phase.

%taint from the address to the code accessing the memory associated with tainted address.
%This allows us to identify what code is accessing what types of data.
%Later allocations do not modify the shadow, unless the thread performing the
%allocation does not 
%When an
%object is allocated, its address is tainted with \TypeId, and, if this is a new \TypeId, then the sha
%and which is initialized 
%Every time a tainted address is accessed, we perform the following: \begin{enumerate}
%\item Propagate 
%
%\item Check if the current thread matches the $AllocThread$. If not, then set $IsShared \gets true$. Future accesses to an
%object with $IsShared = true$ d
%
%\end{enumerate}
% identifier that is
%uniquely associated with the allocation context 
%associated with the allocation context, as represented by a \TypeIdPair{} pair.
%For each dynamically allocated data structure, we ``taint'' the address of the 
%Pruning code 
%This analysis determines what types of data are shared, i.e. accessed by more than one thread, so that the 
%\Toolname{}'s training phase analysis data access patterns to determine what types of data are accessed by more than one
%thread. The insight is that we should only look
%\paragraph{Code Polymorphism Analysis}

% --------------------------------------
\paragraph{Detection}

The detection phase is composed of three components:

1. {\bf The interposition layer}. We interpose on dynamic memory allocators and record the most recently allocated
address for each \TypeId{}. The set of all most recently allocated objects forms the population of addresses from which
we sample.

As a minor optimization, we restrict ourselves to only those {\TypeId}s associated that were found during training
to be associated with shared data structures. Also, unlike during the training phase, we \emph{do not} taint any
addresses.

2.  {\bf The monitor}. The monitor is a separate thread, spawned by \Toolname{} at program startup, and is responsible
for implementing \Toolname's \emph{comprehensive data sampling}. The monitor thread periodically
wakes up, chooses a sample set using a round-robin algorithm over the population of recently allocated addresses by type, and
then enables software watchpoints on all sampled addresses.

After enabling the watchpoints, the monitor goes to sleep. When it next wakes up, it removes any active watchpoints
and generates reports on any cache line contention that was detected. Finally, it restarts the sampling process.

3. {\bf The instrumentation}. There are two tiers of instrumentation: lightweight and heavyweight. Lightweight
instrumentation applies to all basic blocks of code identified during training as \emph{not} containing accesses to shared
data. Lightweight instrumentation exists to maintain Granary's control over the program binary, and therefore to maintain
the comprehensiveness of the analysis.

The heavyweight instrumentation implements a form of ``unlimited watchpoints" \cite{UnlimitedWatchpoints}. The
granularity of each watchpoint is a cache line, but can easily be adjusted in order to model systems with smaller or
larger cache line sizes\footnote{The granularity of shadow memory is tunable via a command-line argument}.
In our setup, every 64 bytes of memory has its own shadow memory and watchpoint. Note that
this does not model actual hardware limits on cache sizes and cache contention caused by associativity.

When a memory address is accessed, the instrumentation checks the shadow memory to see if the watchpoint for
the current address is set. If it is, then the instrumentation code further checks to shadow to see if another thread
owns the cache line (in which case contention is detected), or if the current thread can take ownership of the cache line.

@@

%% --------------------------------------
\paragraph{DataCollider}
DataCollider \cite{DataCollider} is a precise, sampling-basbased oned race detector for the WIndows kernel. DataCollider's key
novelties are its extremely low overhead, as well as its generality: it detects data races without any understanding of
a system's underlying synchronization primitives. DataCollider uniformly samples code by introducing breakpoints into the code.
When a breakpoint is hit, a hardware watchpoint is added to data that was about to be accessed by the interrupted instruction.
Then, the interrupted thread is paused for a short period of time to give another thread the opportunity to access the
same data. If the data is accessed by a second thread while the first thread is paused, then the watchpoint will detect the
access, thus allowing DataCollider to detect a race ``in the act.""

While this approach is elegant in its simplicity, it has two major drawbacks. First, its dependence on hardware watchpoints
limits it to being able to detect races on at most four distinct 8-byte memory locations simultaneously \cite{IntelSDManual}.
Second, in some cases, it cannot be generalized to detecting arbitrary cache contention when the one contending access
\emph{happens-before} the other \cite{VectorClocks}. For example, if true/false sharing frequently occurs within critical sections;
however, this would be missed by a naiive generalization of the DataCollider algorithm to this problem.

%% --------------------------------------
%\paragraph{Predator}

% TODO: I should really write about this, but time :-(

@@

%Intel's VTune \cite{IntelVTune} an an imprecise, sampling-based profiler that can detect false sharing using hardware performance
%counters and monitoring events. It provides detailed, architecture-specific performance-tuning suggestions. Unlike \Toolname{},
%VTune can only report individual program counters that general cache coherency events, but cannot correlate threads by showing
%two contending stack traces.

@@

Plastic \cite{Plastic} is a Xen-based dynamic false sharing detector and repairer. Plastic uses hardware performance counters
to detect when cache line eviction events due to contention exceed some threshold, and then turns on page-granularity
ownership tracking using hardware page protection. When Plastic discovers sharing of a page, it turns on heavyweight recording
of memory access instructions. Later, it analyzes this record to determine what--if any--cache lines on the page had contention,
and then uses a limited form of DBT to ``patch" the virtualized execution so as to remap previously contended accesses into
non-contending accesses. Plastic's dependence on virtualization limit its ability to find false sharing in arbitrary, binary device
drivers that cannot be virtualized.

%
%SHERIFF \cite{SHERIFF} is a precise false-sharing detector and repairer that works in a similar way to Plastic. Instead of
%tracking page ownership at the VMM level, it tracks page ownership on a per-application basis by turning threads into
%processes, thus allowing it to bootstrap off of existing OS-provided process memory isolation. SHERIFF detects contention
%using twinning and diffing of shared pages \cite{TreadMarks}, and fixes false-sharing using a combination of delaying and

Eraser \cite{Eraser} is a dynamic race detector that uses shadow memory to track the set of locks that must be held in order
to access shared data. Eraser initially associates the set of all locks with every shared variable. As shared variables are accessed,
their lock sets are refined based on observing the current set of locks held by a particular thread. If a shared variable's lockset
is refined to the empty set, then a data race has potentially been discovered. The benefit of this approach is that it has the
ability to predict data races, even if two accesses were never truly concurrent. The drawback is that a lockset approach is
challenging to generalize: it requires a complete understanding of a program's locking discipline.

Other work based on Lamport's \emph{happens-before} relationship \cite{VectorClocks} can precisely distinguish data races
from true sharing by recognizing that synchronizing operations (such as locks and condition variables) introduce a strict ordering of
accesses. Under this lens, true sharing occurs when two threads access the same variable, but where one of the accesses happens-%
before the other due to the presence of a synchronizing operation. A data race occurs where neither operation logically happens-%
before the other. Like Eraser, generalizing happens-before to arbitrary COTS binaries can be challenging as it requires an
understanding of the sycnrhonization primitives used by a program.

%Initially,
%it applies read-only hardware page protection to all shared memory. Then, as threads attempt to write to protected pages, those
%pages are privately mapped as read/write. When the same page is privately mapped by two or more threads, SHERIFF applies
%the twinning and diffing approach of TreadMarks  to detect local changes to shared pages and narrow down
%on instances of false sharing. Like \Toolname{}, SHERIFF uses a coarse-grained mechanism to hint at the potential for races.

%
%Like \Toolname{}, it takes a pipeline-based
%approach that successively narrows down on the root causes of false sharing. Plastic also uses DBT as part of its
%repair mechanism. Plastic's pipeline begins by using hardware performance counters related to cache contention
%events to detect the presence of false-sharing. Then, it applies a similar technique to MultiRace \cite{MultiRace}, in
%that it tracks the dominant access pattern by each thread to memory on a shared page. This information informs
%their repair mechanism, which transparently re-maps memory accesses at a byte granularity.

%
%SHERIFF \cite{SHERIFF} is a false-sharing detector and repairer that works in a similar way to MultiRace and Plastic. Initially,
%it applies read-only hardware page protection to all shared memory. Then, as threads attempt to write to protected pages, those
%pages are privately mapped as read/write. When the same page is privately mapped by two or more threads, SHERIFF applies
%the twinning and diffing approach of TreadMarks \cite{TreadMarks} to detect local changes to shared pages and narrow down
%on instances of false sharing. Like \Toolname{}, SHERIFF uses a coarse-grained mechanism to hint at the potential for races.

@@

%Under this categorization scheme, \Toolname{} is a dynamic, precise, variable-granularity, mixed-mode
%false-sharing detector. It draws inspiration from a number of existing race detectors, including LiteRace
%\cite{LiteRace}, MultiRace \cite{MultiRace}, RaceMob \cite{RaceMob}, RaceTrack \cite{RaceTrack}, and
%DataCollider \cite{DataCollider}.
%
%LiteRace \cite{LiteRace} is a precise, sampling-based dynamic race detector implemented on top of the Microsoft
%Pheonix binary translator. It samples at a function granularity, and executes either one specialized version of the
%code or another. The sampling rate for a function is inversely proportional to how often it executes. This is based
%on the insight that races are likely to occur in infrequently executed code, because otherwise they are benign or
%would have been noticed and resolved. This insight, called the ``cold code hypothesis", motivated \Toolname's
%post-mortem ranking scheme: cases of false sharing are more highly ranked when one---but \emph{not} both---of
%the participating threads are executing cold code. This variant on the cold code hypothesis emphasizes the similarly
%hard-to-find case of cold code pessimizing hot code.

%
%MultiRace \cite{MultiRace} is a comprehensive, dynamic race detector that employs a hybrid happens-before and
%lockset-based approach. It operates at an object granularity, and therefore reports false-positives. It is unique
%in that it uses ``pointer swizzing" as a means of remapping memory to ``minipages" that track the dominating
%access pattern of a thread (read-write > read-only > no-access). \Toolname{} uses a similar pointer swizzing approach, but
%in the form of address watchpoints \cite{AddressWatchpoints}. Pointers to heap-allocated objects are ``swizzled" to
%include a type identifier that is unique to a \TypeIdPair{} pair. This improves the diagnosis of false-sharing,
%because it allows \Toolname{} to distinguish between two threads accessing different fields of the same object, and
%two threads accessing two nearby objects of the same (or different) types.
%
%RaceMob \cite{RaceMob} and RaceTrack \cite{RaceTrack} are two hybrid happens-before and lockset-based race detectors.
%RaceMob begins by statically analyzing source code to find potential races (lockset), then dynamically verifies those races
%(happens-before). RaceTrack is a JIT-based race-detector for Microsoft's common language runtime (CLR). It uses the
%concept of threadsets to efficiently maintain happens-before relationships for memory locations and detect the possible
%presence of a race and uses lockets to confirm predicted races. The novelty of RaceTrack is that it adapts the detection
%granularity from object-granularity to field-granularity. \Toolname{} follows the same high-level approach as both RaceMob
%and RaceTrack in that it attempts to predict the source of false-sharing at one granularity, then verify those predictions 
%at another granularity.

@@

%\paragraph{Sampling-based}
%\paragraph{Comprehensive}

@@

\Toolname{} can be thought of as combining DataCollider's sampling based approach \cite{DataCollider}, and Dr. Memory's
cache line ownership tracking approach \cite{DrContention}. The major distinction between \Toolname, DataCollider, and
Dr. Memory are that:
\begin{itemize}
\item \Toolname's sampling approach is data-centric, as opposed to DataCollider's, which is code-centric. That is, DataCollider
interposes on code first in order to find data that should be watched. \Toolname{} starts with the data that it expects to
be shared (based on training), and then waits for code to access that data.

\item Unlike Dr. Memory, \Toolname{} does not heavily instrument all code for cache line ownership tracking, nor does it
persistently track the ownership of all cache lines. Instead, the training phase informs the detection phase on what
instructions to instrument, and sampling turns on temporary tracking of a subset of cache lines. When the \emph{first}
instance of contention is detected, then the issue is reported and tracking is disabled.
\end{itemize}

% --------------------------------------
\subsection{Ownership Tracking}
\begin{figure}
\begin{center}
\epsfig{file=states.eps, height=30em}
\end{center}
\caption{\label{fig:states_of_shadow}States of individual shadow memory cells. Each cell tracks the watched and ownership %
status of a cache line. Yellow states are unwatched. Green states are watched. Red states represent contention that was missed %
due to races in the instrumentation. Transitions between states are based on (potentially stale) observed prior states. For %
example, the spurious acquire transition is possible because the decision to make the transition is based on a racy \texttt{CMP} %
against the $SampleId$ (\figref{inline_assembly}). If two threads simultaneously see that the cache line it watched but not %
owned, then both will execute the atomic \texttt{XCHG}, and only one thread will ``win". The ignore transition is induced, %
either because the sampling timeout expires, or more likely, because the report transition occurs.}
\end{figure}

\begin{figure}
\lstset{language=[x64]Assembler}
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
;; Racy watchpoint check.
CMP [ShadowAddr + Offs(SampleId)], 0
JZ ignore
;; Racy ownership check.
CMP [ShadowAddr + Offs(ThreadId)], CurrThreadId
JZ ignore
;; Take ownership, or transfer ownership
LOCK XCHG [ShadowAddr], OldState
CALL CheckContention(OldState,Addr,PC)
ignore:
\end{lstlisting}
\caption{\label{fig:inline_assembly}Instrumentation instructions injected for each shared memory access. The algorithm optimizes on %
instruction footprint by using a single \texttt{XCHG} instruction to perform multiple state transitions from %
\Figref{states_of_shadow}, and then figuring out what state we should be in based on the value of $OldState$.}
\end{figure}
